Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-01-21 pkg/disttree/inst/draft/draft.Rnw
small corrections and qqr-plot now in appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1967 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 25666be79535b36b256c592f8f05a8a432ed2147

Forest models differ mainly in the way how the trees are combined to make predictions. Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{z}$ and the resulting subgroups and weighing schemes. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process. In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations. This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``random survival forests'' by \cite{ishwaran2008}. The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (“bagging survival trees”, Hothorn et al. 2004). 
In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
$$
w^{\text{forest}}_i(\bold{z}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.

@@

In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std. dev., ensemble minimum and ensemble maximum. All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std. dev. of mean/minimal/maximal values respectively. Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.

@@

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. dev. of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. dev. of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. dev. of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. dev. of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. dev. of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}

@@

%\begin{figure}[t!]
%%\begin{subfigure}{0.5\textwidth}
%<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE>>=
%    
%# out of sample
%set.seed(7)
%par(mfrow = c(2,2))
%set.seed(1)
%qqrplot(pit_df_t, nsim = 1, main = "distforest (out of bag)")
%qqrplot(pit_g_t, nsim = 1, main = "gamlss (out of bag)")
%qqrplot(pit_gb_t, nsim = 1, main = "gamboostLSS (out of bag)")
%qqrplot(pit_ml_t, nsim = 1, main = "EMOS (out of bag)")

%@
%\caption{\label{qqr_oosample} QQR-Plot for station Axams (out of sample)}
%\end{figure}

@@

%\newpage
%\section*{Appendix}

>>>>

Forest models differ mainly in the way how the trees are combined to make predictions. Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{z}$ and the resulting subgroups and weighing schemes. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process. In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations. This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``random survival forests'' by \cite{ishwaran2008}. The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (“bagging survival trees”, Hothorn et al. 2004). 
In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
$$
w^{\text{forest}}_i(\bold{z}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.

@@

In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std. deviation, ensemble minimum and ensemble maximum. All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std. deviation of mean/minimal/maximal values respectively. Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.

@@

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}

@@

@@

%\newpage
\section*{Appendix}

\begin{figure}[t!]
\begin{center}
%\begin{subfigure}{0.8\textwidth}
<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=

# out of sample
set.seed(7)
par(mfrow = c(2,2))
set.seed(1)
qqrplot(pit_df_t, nsim = 1, main = "distributional forest", ylim = c(-5,5))
qqrplot(pit_g_t, nsim = 1, main = "prespecified GAM", ylim = c(-5,5))
qqrplot(pit_gb_t, nsim = 1, main = "boosted GAM", ylim = c(-5,5))
qqrplot(pit_ml_t, nsim = 1, main = "EMOS", ylim = c(-5,5))

@
\caption{\label{qqr_oosample} Out-of-sample QQR-plot for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
\end{center}
\end{figure}
